---
title: "BCRS-28 | Script 1: scRNA-seq Analysis"
subtitle: "Cell Annotation, Differential Expression & Pathway Enrichment"
author: "Zinab O. Doha"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)
```

---

## Overview

This script performs single-cell RNA sequencing analysis of BRCA1-mutant
versus wild-type normal breast epithelium, including:

- Quality control and normalization
- Dimensionality reduction and clustering
- Cell type annotation using canonical markers
- Epithelial subset differential expression (BRCA1 vs WT)
- Pathway enrichment analysis (GO, KEGG, GSEA Hallmark)
- Figure 1 (cell atlas) and Figure 2 (epithelial DE) generation

**Input:** `seu_wt_brca1_with_metadata.rds`  
**Output:** Figures 1–2, Supplementary Tables S1–S4

---

## Step 1 — Libraries

```{r libraries}
suppressPackageStartupMessages({
  library(Seurat)
  library(tidyverse)
  library(patchwork)
  library(pheatmap)
  library(ggrepel)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(enrichplot)
  library(msigdbr)
  library(openxlsx)
  library(DESeq2)
  library(tibble)
  library(dplyr)
})

dir.create("figures",       showWarnings = FALSE)
dir.create("supplementary", showWarnings = FALSE)
set.seed(42)

# ── Color palette ────────────────────────────────────────────────────────────
COL_BRCA1 <- "#E41A1C"   # Red
COL_WT    <- "#2ABAB4"   # Teal

cat("Libraries loaded.\n")
```

---

## Step 2 — Load & Inspect Object

```{r load}
seu <- readRDS("seu_wt_brca1_with_metadata.rds")

cat("=== Object summary ===\n")
print(seu)
cat("\nGenotype counts:\n")
print(table(seu$genotype))
cat("\nCell type counts:\n")
print(table(seu$CellType))
```

---

## Step 3 — Normalize & Dimensionality Reduction

```{r reprocess}
DefaultAssay(seu) <- "RNA"

seu <- NormalizeData(seu, normalization.method = "LogNormalize",
                     scale.factor = 10000, verbose = FALSE)
seu <- FindVariableFeatures(seu, selection.method = "vst",
                             nfeatures = 3000, verbose = FALSE)
seu <- ScaleData(seu, features = VariableFeatures(seu), verbose = FALSE)
gc()

seu <- RunPCA(seu,         npcs = 30,       verbose = FALSE)
seu <- FindNeighbors(seu,  dims = 1:20,     verbose = FALSE)
seu <- FindClusters(seu,   resolution = 0.25, verbose = FALSE)
seu <- RunUMAP(seu,        dims = 1:20, seed.use = 42, verbose = FALSE)

cat("Clusters found:", length(unique(seu$seurat_clusters)), "\n")
```

---

## Step 4 — Cell Type Annotation {.tabset}

### Canonical Marker Dotplot

```{r canonical-markers, fig.width=14, fig.height=6}
canonical_markers <- intersect(
  c("PTPRC", "CD79A", "MS4A1",
    "CD14",  "CD68",  "LYZ",
    "CD3D",  "CD3E",  "CD8A",
    "EPCAM", "KRT8",  "KRT18",
    "COL1A1","FAP",
    "PECAM1","CDH5",  "CLDN5"),
  rownames(seu)
)

Idents(seu) <- "seurat_clusters"

DotPlot(seu,
        features  = canonical_markers,
        dot.scale = 6,
        col.min   = 0) +
  scale_colour_gradient2(low="white", mid="#FB9A99",
                         high="#E31A1C", midpoint=1) +
  theme(axis.text.x = element_text(angle=45, hjust=1, size=10)) +
  labs(title="Canonical markers per cluster")
```

### Cluster Assignment

```{r annotation}
cluster_to_celltype <- c(
  "0"  = "Epithelial", "1"  = "Fibroblast",  "2"  = "Endothelial",
  "3"  = "Epithelial", "4"  = "Epithelial",  "5"  = "Epithelial",
  "6"  = "Epithelial", "7"  = "Epithelial",  "8"  = "Epithelial",
  "9"  = "T_cells",    "10" = "Myeloid",      "11" = "Epithelial",
  "12" = "Epithelial", "13" = "Myeloid",      "14" = "B_cells",
  "15" = "Epithelial", "16" = "Endothelial"
)

ct_vec        <- cluster_to_celltype[as.character(seu$seurat_clusters)]
names(ct_vec) <- colnames(seu)
seu           <- AddMetaData(seu, metadata = ct_vec,
                              col.name = "CellType_manual")

broad_vec <- case_when(
  ct_vec %in% c("T_cells","Myeloid","B_cells") ~ "Immune",
  ct_vec == "Epithelial"                        ~ "Epithelial",
  ct_vec %in% c("Fibroblast","Endothelial")     ~ "Stromal"
)
names(broad_vec) <- colnames(seu)
seu <- AddMetaData(seu, metadata = broad_vec, col.name = "BroadType")

cat("Cell type distribution:\n")
print(table(seu$CellType_manual))
cat("\nGenotype x CellType:\n")
print(table(seu$genotype, seu$CellType_manual))
```

---

## Step 5 — Figure 1: Single-Cell Atlas {.tabset}

### Panel A — Clusters

```{r fig1a}
fig1A <- DimPlot(seu, group.by="seurat_clusters",
                 label=TRUE, pt.size=0.3) +
  NoLegend() + labs(title="Unsupervised clusters")
fig1A
```

### Panel B — Cell Types

```{r fig1b}
fig1B <- DimPlot(seu, group.by="CellType_manual",
                 label=TRUE, repel=TRUE, pt.size=0.3) +
  NoLegend() + labs(title="Cell type annotation")
fig1B
```

### Panel C — Donor

```{r fig1c}
fig1C <- DimPlot(seu, group.by="orig.ident", pt.size=0.3) +
  labs(title="Sample / donor")
fig1C
```

### Panel D — Marker Dotplot

```{r fig1d, fig.width=12, fig.height=5}
Idents(seu) <- "CellType_manual"
markers_dot <- intersect(
  c("PTPRC","CD79A","MS4A1","CD14","CD68","LYZ",
    "CD3D","CD3E","CD8A","EPCAM","KRT8","KRT18",
    "COL1A1","FAP","PECAM1","CDH5","CLDN5"),
  rownames(seu)
)

fig1D <- DotPlot(seu, features=markers_dot, dot.scale=6, col.min=0) +
  scale_colour_gradient2(low="white", mid="#FB9A99",
                         high="#E31A1C", midpoint=1) +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  labs(title="Canonical markers", x=NULL, y=NULL)
fig1D
```

### Compose & Save

```{r fig1-save, fig.width=18, fig.height=13}
fig1_final <- ((fig1A | fig1B | fig1C) +
                 plot_layout(widths=c(1,1,1.15))) /
              fig1D +
  plot_layout(heights=c(1.2,1)) +
  plot_annotation(
    title    = "Figure 1 · scRNA-seq Landscape of BRCA1 vs WT Breast Tissue",
    subtitle = paste0(ncol(seu), " cells | Unsupervised clustering + manual annotation"),
    theme    = theme(
      plot.title    = element_text(size=16, face="bold", hjust=0.5),
      plot.subtitle = element_text(size=12, colour="grey40", hjust=0.5)
    )
  )

fig1_final

ggsave("figures/Figure1_atlas.png",  fig1_final,
       width=18, height=13, dpi=300)
ggsave("figures/Figure1_atlas.pdf",  fig1_final,
       width=18, height=13)
cat("Figure 1 saved.\n")
```

---

## Step 6 — Subset Epithelial Cells

```{r subset-epi}
seu_epi <- subset(seu, subset = BroadType == "Epithelial")
seu_epi$genotype <- factor(seu_epi$genotype, levels=c("WT","BRCA1"))

cat("Epithelial cells:", ncol(seu_epi), "\n")
print(table(seu_epi$genotype))

DefaultAssay(seu_epi) <- "RNA"
seu_epi <- NormalizeData(seu_epi, verbose=FALSE)
seu_epi <- FindVariableFeatures(seu_epi, nfeatures=3000, verbose=FALSE)
seu_epi <- ScaleData(seu_epi,
                     features=VariableFeatures(seu_epi), verbose=FALSE)
gc()
seu_epi <- RunPCA(seu_epi,       npcs=30,         verbose=FALSE)
seu_epi <- FindNeighbors(seu_epi, dims=1:20,      verbose=FALSE)
seu_epi <- FindClusters(seu_epi,  resolution=0.4, verbose=FALSE)
seu_epi <- RunUMAP(seu_epi,       dims=1:20, seed.use=42, verbose=FALSE)

cat("Epithelial subset processed.\n")
```

---

## Step 7 — Differential Expression (BRCA1 vs WT) {.tabset}

### Run FindMarkers

```{r de-analysis}
Idents(seu_epi) <- "genotype"

de_results <- FindMarkers(
  seu_epi,
  ident.1         = "BRCA1",
  ident.2         = "WT",
  test.use        = "wilcox",
  min.pct         = 0.1,
  logfc.threshold = 0.1,
  verbose         = FALSE
) %>%
  rownames_to_column("gene") %>%
  rename(log2FoldChange = avg_log2FC,
         padj           = p_val_adj,
         pvalue         = p_val) %>%
  arrange(padj, desc(abs(log2FoldChange)))

cat("Up in BRCA1 (FDR<0.05, |LFC|>0.5):",
    sum(de_results$padj < 0.05 & de_results$log2FoldChange > 0.5), "\n")
cat("Down in BRCA1:",
    sum(de_results$padj < 0.05 & de_results$log2FoldChange < -0.5), "\n")
```

### Pseudobulk Validation (DESeq2)

```{r pseudobulk}
# ── Defined per reviewer comment 2 ──────────────────────────────────────────
DONOR_COL    <- "donor_id"
GENOTYPE_COL <- "genotype"

BCRS28_genes <- c(
  "GADD45G",
  "CRIP1",   "SNCG",    "PIP",
  "KRT14",   "KRT15",   "MUCL1",   "SERHL2",
  "ANGPTL4", "BAMBI",   "CD14",    "CXCL14",
  "KLK7",    "KLK5",    "RARRES1", "SLC40A1", "FABP3",
  "TGM2",    "HMOX1",   "SCGB3A1", "SERPINA5","FASN",
  "WNT4",    "LMO7",    "CCL20",   "EEF1A2",  "CCDC74A", "GBP1"
)

# Aggregate counts per donor
DefaultAssay(seu_epi) <- "RNA"
counts    <- GetAssayData(seu_epi, layer="counts")
meta_epi  <- seu_epi@meta.data
donors    <- unique(meta_epi[[DONOR_COL]])

pb_counts <- sapply(donors, function(d) {
  cells <- rownames(meta_epi)[meta_epi[[DONOR_COL]] == d]
  if (length(cells)==1) as.numeric(counts[,cells])
  else                   rowSums(counts[,cells])
})
pb_counts           <- as.matrix(pb_counts)
colnames(pb_counts) <- donors
rownames(pb_counts) <- rownames(counts)

# Filter low counts
keep      <- rowSums(pb_counts >= 10) >= 2
pb_counts <- pb_counts[keep,]

# Build coldata
coldata <- meta_epi %>%
  dplyr::select(all_of(c(DONOR_COL, GENOTYPE_COL))) %>%
  distinct()
rownames(coldata) <- coldata[[DONOR_COL]]
coldata[[DONOR_COL]] <- NULL
coldata <- coldata[colnames(pb_counts),,drop=FALSE]
coldata[[GENOTYPE_COL]] <- factor(coldata[[GENOTYPE_COL]],
                                   levels=c("WT","BRCA1"))

# DESeq2
dds <- DESeqDataSetFromMatrix(
  countData = pb_counts,
  colData   = coldata,
  design    = ~ genotype
)
dds <- DESeq(dds)
res <- results(dds, contrast=c("genotype","BRCA1","WT"), alpha=0.05)

cat("\nPseudobulk DESeq2 summary:\n")
summary(res)

# BCRS-28 pseudobulk results
res_pb <- as.data.frame(res) %>%
  rownames_to_column("gene") %>%
  filter(gene %in% BCRS28_genes) %>%
  dplyr::select(gene, log2FoldChange, pvalue, padj) %>%
  mutate(
    log2FoldChange = round(log2FoldChange, 3),
    padj           = signif(padj, 3),
    Direction      = ifelse(log2FoldChange > 0, "UP", "DOWN"),
    Significant    = ifelse(!is.na(padj) & padj < 0.05, "YES", "NO")
  ) %>%
  arrange(padj)

cat("\nBCRS-28 pseudobulk results:\n")
print(res_pb)

write.csv(res_pb, "supplementary/TableS7_BCRS28_Pseudobulk.csv",
          row.names=FALSE)
cat("✓ Table S7 saved.\n")
```

### Donor Consistency

```{r donor-consistency}
bcrs_present <- BCRS28_genes[BCRS28_genes %in% rownames(seu_epi)]

expr_data <- GetAssayData(seu_epi, assay="RNA", layer="data")
expr_bcrs <- expr_data[bcrs_present,]

donor_expr <- sapply(donors, function(d) {
  cells <- rownames(meta_epi)[meta_epi[[DONOR_COL]] == d]
  if (length(cells)==1) as.numeric(expr_bcrs[,cells])
  else                   rowMeans(expr_bcrs[,cells])
})
donor_expr           <- as.data.frame(donor_expr)
colnames(donor_expr) <- donors
rownames(donor_expr) <- bcrs_present

# Annotation
col_annot <- meta_epi %>%
  dplyr::select(all_of(c(DONOR_COL, GENOTYPE_COL))) %>%
  distinct()
rownames(col_annot) <- col_annot[[DONOR_COL]]
col_annot[[DONOR_COL]] <- NULL
col_annot <- col_annot[colnames(donor_expr),,drop=FALSE]
colnames(col_annot) <- "Genotype"

# Heatmap
pheatmap(
  donor_expr,
  annotation_col    = col_annot,
  annotation_colors = list(Genotype=c(BRCA1=COL_BRCA1, WT=COL_WT)),
  cluster_rows      = TRUE,
  cluster_cols      = TRUE,
  scale             = "row",
  color             = colorRampPalette(c(COL_WT,"white",COL_BRCA1))(100),
  fontsize_row      = 9,
  fontsize_col      = 8,
  main              = "BCRS-28 Expression Across Individual Donors\n(Supplementary Figure S1D)",
  border_color      = NA,
  filename          = "figures/FigS1D_BCRS28_Donor_Heatmap.png",
  width=9, height=8
)
cat("✓ Figure S1D saved.\n")

# Consistency table
brca1_donors <- rownames(col_annot)[col_annot$Genotype == "BRCA1"]
wt_donors    <- rownames(col_annot)[col_annot$Genotype == "WT"]

consistency_df <- lapply(bcrs_present, function(g) {
  b_vals  <- as.numeric(donor_expr[g, brca1_donors])
  wt_mean <- mean(as.numeric(donor_expr[g, wt_donors]), na.rm=TRUE)
  b_mean  <- mean(b_vals, na.rm=TRUE)
  pct     <- round(mean(b_vals > wt_mean, na.rm=TRUE)*100, 1)
  data.frame(
    Gene                  = g,
    Direction             = ifelse(b_mean > wt_mean, "UP", "DOWN"),
    Pct_Donors_Consistent = pct,
    BRCA1_Mean            = round(b_mean,   3),
    WT_Mean               = round(wt_mean,  3)
  )
}) %>% bind_rows() %>% arrange(desc(Pct_Donors_Consistent))

print(consistency_df)
write.csv(consistency_df,
          "supplementary/TableS8_Donor_Consistency.csv",
          row.names=FALSE)
cat("✓ Table S8 saved.\n")
```

---

## Step 8 — Figure 2: Epithelial DE Visualization {.tabset}

### Panel A — UMAP by Genotype

```{r fig2a}
fig2A <- DimPlot(seu_epi,
                 group.by = "genotype",
                 cols     = c(BRCA1=COL_BRCA1, WT=COL_WT),
                 pt.size  = 0.5) +
  labs(title = "Epithelial cells by genotype")
fig2A
```

### Panel B — Volcano Plot

```{r fig2b, fig.width=9, fig.height=7}
LFC_THRESH <- 0.5
P_THRESH   <- 0.05

top_up   <- de_results %>%
  filter(padj<P_THRESH, log2FoldChange>LFC_THRESH) %>%
  slice_max(log2FoldChange, n=15) %>% pull(gene)
top_down <- de_results %>%
  filter(padj<P_THRESH, log2FoldChange< -LFC_THRESH) %>%
  slice_min(log2FoldChange, n=15) %>% pull(gene)

vol_df <- de_results %>%
  mutate(
    sig = case_when(
      padj < P_THRESH & log2FoldChange >  LFC_THRESH ~ "Up in BRCA1",
      padj < P_THRESH & log2FoldChange < -LFC_THRESH ~ "Down in BRCA1",
      TRUE ~ "NS"
    ),
    label = case_when(
      gene %in% top_up   ~ gene,
      gene %in% top_down ~ gene,
      TRUE ~ NA_character_
    )
  )

n_up   <- sum(vol_df$sig=="Up in BRCA1")
n_down <- sum(vol_df$sig=="Down in BRCA1")
n_ns   <- sum(vol_df$sig=="NS")

fig2B <- ggplot(vol_df, aes(x=log2FoldChange,
                              y=-log10(pvalue+1e-300),
                              colour=sig)) +
  geom_point(size=0.8, alpha=0.6) +
  geom_text_repel(aes(label=label), size=4, fontface="italic",
                  na.rm=TRUE, max.overlaps=20, colour="black") +
  geom_vline(xintercept=c(-LFC_THRESH, LFC_THRESH),
             linetype="dashed", colour="grey40") +
  geom_hline(yintercept=-log10(P_THRESH),
             linetype="dashed", colour="grey40") +
  scale_colour_manual(
    values = c("Up in BRCA1"=COL_BRCA1, "Down in BRCA1"=COL_WT,
               "NS"="grey75"),
    labels = c(paste0("Up in BRCA1 (n=",   n_up,   ")"),
               paste0("Down in BRCA1 (n=", n_down, ")"),
               paste0("NS (n=",            n_ns,   ")")),
    name   = NULL
  ) +
  labs(title    = "Differential Expression — BRCA1 vs WT Epithelial Cells",
       subtitle = paste0("scRNA-seq | Wilcoxon test | FC threshold = ",
                         LFC_THRESH, " | FDR < ", P_THRESH),
       x = "log\u2082 Fold Change (BRCA1 / WT)",
       y = expression(-log[10](p~value))) +
  theme_classic(base_size=12) +
  theme(plot.title    = element_text(face="bold", hjust=0.5),
        plot.subtitle = element_text(hjust=0.5, colour="grey50"),
        legend.position = "top")

fig2B
ggsave("figures/Figure2B_volcano.png", fig2B, width=9, height=7, dpi=300)
cat("Figure 2B saved.\n")
```

### Panel C — Heatmap

```{r fig2c, fig.width=10, fig.height=8}
# Gene lists from verified supplementary table
top_up_hm <- c("SLC40A1","FABP3","CRIP1","SERHL2","RARRES1","CAVIN3",
                "CD14","KLK7","BAMBI","KRT14","PIP","GADD45G","KLK5",
                "MUCL1","SNCG","ANGPTL4","KRT15","CXCL14","ZBTB16",
                "RGS2","ADRB2","DNAJB1","PEG10","SETD9","TOB1")
top_dn_hm <- c("IGHA2","TGM2","WNT4","SCGB3A1","LMO7","CCL20","EEF1A2",
               "SERPINA5","ZNF385C","FASN","HMOX1","CCDC74A","GBP1",
               "CTXN1","MYEOV","AP000851.1","SDCBP2","MMP7","SLC7A2",
               "PRKAA2","SYT7","OPN3","BATF","RAB31","TRIO")

top_up_hm <- intersect(top_up_hm, rownames(seu_epi))
top_dn_hm <- intersect(top_dn_hm, rownames(seu_epi))
hm_genes  <- c(top_up_hm, top_dn_hm)

set.seed(42)
c_brca <- sample(WhichCells(seu_epi, expression = genotype=="BRCA1"),
                 min(15000, sum(seu_epi$genotype=="BRCA1")))
c_wt   <- sample(WhichCells(seu_epi, expression = genotype=="WT"),
                 min(15000, sum(seu_epi$genotype=="WT")))
cells_ord <- c(c_brca, c_wt)

seu_sub <- seu_epi[hm_genes, cells_ord]
seu_sub <- ScaleData(seu_sub, features=hm_genes, verbose=FALSE)
expr_hm <- as.matrix(GetAssayData(seu_sub, layer="scale.data")[hm_genes,])
expr_hm <- pmin(pmax(expr_hm, -2), 2)

ann_col <- data.frame(
  Genotype  = factor(seu_epi$genotype[cells_ord], levels=c("BRCA1","WT")),
  row.names = cells_ord
)
ann_colors <- list(Genotype=c(BRCA1=COL_BRCA1, WT=COL_WT))

png("figures/Figure2C_heatmap.png", width=10, height=8,
    units="in", res=300)
pheatmap(expr_hm,
         color             = colorRampPalette(c("#4575B4","white","#D73027"))(200),
         breaks            = seq(-2, 2, length.out=201),
         cluster_rows      = FALSE,
         cluster_cols      = FALSE,
         show_colnames     = FALSE,
         show_rownames     = TRUE,
         fontsize_row      = 8,
         annotation_col    = ann_col,
         annotation_colors = ann_colors,
         gaps_col          = length(c_brca),
         gaps_row          = length(top_up_hm),
         main              = "Top 25 up & down — BRCA1 vs WT",
         border_color      = NA,
         treeheight_col    = 0,
         treeheight_row    = 0)
dev.off()
cat("Figure 2C saved.\n")
```

---

## Step 9 — Pathway Enrichment {.tabset}

### GO Enrichment (Up genes)

```{r go-up}
up_genes   <- de_results %>%
  filter(padj<0.05, log2FoldChange>0) %>% pull(gene)
up_entrez  <- bitr(up_genes, "SYMBOL","ENTREZID", org.Hs.eg.db)

go_up <- enrichGO(gene          = up_entrez$ENTREZID,
                  OrgDb         = org.Hs.eg.db,
                  ont           = "BP",
                  pAdjustMethod = "BH",
                  pvalueCutoff  = 0.05,
                  readable      = TRUE)

dotplot(go_up, showCategory=15,
        title="GO BP — Up-regulated in BRCA1") +
  theme(plot.title=element_text(face="bold", hjust=0.5))

write.csv(as.data.frame(go_up),
          "supplementary/TableS_GO_UP_BRCA1.csv", row.names=FALSE)
```

### GO Enrichment (Down genes)

```{r go-down}
dn_genes  <- de_results %>%
  filter(padj<0.05, log2FoldChange<0) %>% pull(gene)
dn_entrez <- bitr(dn_genes, "SYMBOL","ENTREZID", org.Hs.eg.db)

go_dn <- enrichGO(gene          = dn_entrez$ENTREZID,
                  OrgDb         = org.Hs.eg.db,
                  ont           = "BP",
                  pAdjustMethod = "BH",
                  pvalueCutoff  = 0.05,
                  readable      = TRUE)

dotplot(go_dn, showCategory=15,
        title="GO BP — Down-regulated in BRCA1") +
  theme(plot.title=element_text(face="bold", hjust=0.5))

write.csv(as.data.frame(go_dn),
          "supplementary/TableS_GO_DOWN_BRCA1.csv", row.names=FALSE)
```

### GSEA Hallmark

```{r gsea}
ranked_vec <- setNames(
  sign(de_results$log2FoldChange) * (-log10(de_results$pvalue+1e-300)),
  de_results$gene
)
ranked_vec <- sort(ranked_vec, decreasing=TRUE)
ranked_vec <- ranked_vec[!duplicated(names(ranked_vec))]

gene_map   <- bitr(names(ranked_vec), "SYMBOL","ENTREZID", org.Hs.eg.db)
ranked_e   <- ranked_vec[gene_map$SYMBOL]
names(ranked_e) <- gene_map$ENTREZID
ranked_e   <- sort(ranked_e[!duplicated(names(ranked_e))], decreasing=TRUE)

h_sets <- msigdbr(species="Homo sapiens", category="H") %>%
  dplyr::select(gs_name, entrez_gene) %>%
  mutate(entrez_gene=as.character(entrez_gene))

gsea_h <- GSEA(ranked_e,
               TERM2GENE     = h_sets,
               minGSSize     = 15,
               pvalueCutoff  = 0.05,
               pAdjustMethod = "BH",
               seed          = TRUE,
               eps           = 0,
               verbose       = FALSE)

cat("Significant Hallmark pathways:", nrow(as.data.frame(gsea_h)), "\n")

gsea_df <- as.data.frame(gsea_h) %>%
  mutate(
    pathway   = gsub("HALLMARK_","",ID) %>%
                gsub("_"," ",.) %>% str_to_title() %>% str_wrap(35),
    direction = ifelse(NES>0, "Up in BRCA1", "Down in BRCA1")
  ) %>%
  arrange(NES) %>%
  mutate(pathway=factor(pathway, levels=pathway))

ggplot(gsea_df, aes(x=NES, y=pathway, fill=direction)) +
  geom_bar(stat="identity", width=0.75) +
  geom_vline(xintercept=0, colour="grey30") +
  scale_fill_manual(
    values = c("Up in BRCA1"=COL_BRCA1, "Down in BRCA1"=COL_WT)
  ) +
  labs(title="Hallmark GSEA — BRCA1 vs WT",
       x="Normalized Enrichment Score (NES)", y=NULL) +
  theme_classic(base_size=12) +
  theme(plot.title=element_text(face="bold", hjust=0.5))

ggsave("figures/Figure2_GSEA_hallmark.png", width=9, height=5, dpi=300)
```

---

## Step 10 — Save Supplementary Tables

```{r save-tables}
wb <- createWorkbook()

addWorksheet(wb, "S1_DE_Full")
writeData(wb, "S1_DE_Full", de_results)

addWorksheet(wb, "S2_DE_Significant")
writeData(wb, "S2_DE_Significant",
          de_results %>% filter(padj < 0.05))

addWorksheet(wb, "S3_Top25_Up")
writeData(wb, "S3_Top25_Up",
          de_results %>%
            filter(padj<0.05, log2FoldChange>0) %>%
            slice_max(log2FoldChange, n=25))

addWorksheet(wb, "S4_Top25_Down")
writeData(wb, "S4_Top25_Down",
          de_results %>%
            filter(padj<0.05, log2FoldChange<0) %>%
            slice_min(log2FoldChange, n=25))

saveWorkbook(wb, "supplementary/Tables_S1_S4_DE_Results.xlsx",
             overwrite=TRUE)
cat("✓ Supplementary Tables S1-S4 saved.\n")
```

---

## Session Info

```{r session}
sessionInfo()
```
